Global well-being and mental health in the internet age

### 2023-07-17 ms.Rmd
Move a bunch of stuff to a separate file

---
title: Global well-being and mental health in the internet age
shorttitle: Global well-being
leftheader: Global well-being
author: 
  - name: Matti Vuorre
    affiliation: "1,2"
    address: Tilburg University
    email: m.j.vuorre@tilburguniversity.edu
  - name: Andrew K. Przybylski
    affiliation: "2"
    corresponding: yes
    email: andy.przybylski@oii.ox.ac.uk
affiliation:
  - id: 1
    institution: Tilburg School of Social and Behavioral Sciences, Tilburg University
  - id: 2
    institution: Oxford Internet Institute, University of Oxford
authornote: |
  \noindent \textbf{This pre-print is not yet peer-reviewed.}
abstract: | 
  In the last two decades the rise of internet technologies have impacted nearly all domains of human life and inspired concern that their widespread adoption has impacted mental health and psychological well-being. However, research on the topic is contested and hampered by methodological shortcomings leaving the broader consequences of internet adoption unknown. We show that the past two decades have seen small and inconsistent changes in global well-being and mental health. The broadest available health and internet technology data does not support the idea that the adoption of internet and mobile broadband is consistently linked to negative psychological outcomes. However, associations between internet technology adoption and depression and self-harm were strongest (albeit still small) for individuals aged between 10 and 34. Further investigation of this topic requires transparent study of online behaviours where they occur, namely on online platforms. We call for increased collaborative efforts between independent scientists and the internet technology sector.

keywords: well-being, mental health, internet technology, technology effects
wordcount: '`r scales::number(wordcountaddin:::word_count("ms.Rmd"), big.mark = ",")`'
bibliography: "references.bib"
floatsintext: yes
linenumbers: no
draft: no
mask: no
figurelist: no
tablelist: no
footnotelist: no
csl: "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
documentclass: apa7
classoption: jou
output: 
  papaja::apa6_pdf:
    number_sections: false
    keep_tex: false
  papaja::apa6_docx: 
    number_sections: false
editor_options: 
  chunk_output_type: console
always_allow_html: true
---

```{r setup, include = FALSE}
# Packages
library(papaja)
library(readxl)
library(imputeTS)
library(scales)
library(janitor)
library(haven)
library(labelled)
library(tidybayes)
library(ggdist)
library(countrycode)
library(lubridate)
library(brms)
library(patchwork)
library(knitr)
library(ggstance)
library(bayestestR)
library(ggh4x)
library(tidyverse)

opts_chunk$set(
  eval = TRUE,
  cache = TRUE,
  include = FALSE,
  message = FALSE,
  warning = FALSE,
  dpi = 300,
  fig.align = "center"
)
# For tables: Show NAs as blanks
options(knitr.kable.NA = '')

# Create directory for intermediate files
dir.create("data", FALSE)

# Plotting options
theme_set(
  theme_linedraw(
    base_size = if_else(knitr::is_latex_output(), 9, 8)
  ) +
    theme(
      panel.grid = element_blank(),
      strip.text = element_text(margin = margin(4, 4, 4, 4, "pt")),
      axis.text = element_text(size = rel(0.75)),
      axis.ticks = element_line(size = rel(0.3))
    )
)
```

@@

```{r data-clean}
## # Understand variables & values
## # Codes NOTE 1: yes 2: no!
## read_spss(
##   "data-raw/Gallup/The_Gallup_101521.sav",
##   n_max = 1,
##   col_select = c(
##     WPID, WP1219,
##     WP16,
##     WP60, WP61, WP63, WP65, WP67,
##     WP68, WP69, WP70, WP71, WP74
##   )
## ) %>%
##   generate_dictionary()

# First liberate the data from the slow SPSS file and save to disk
gwp_path <- "data-raw/Gallup/gwp-processed.rds"
if (!file.exists(gwp_path)) {
  gwp <- read_spss(
    "data-raw/Gallup/The_Gallup_101521.sav",
    col_select = c(
      YEAR_CALENDAR,
      COUNTRYNEW,
      WPID, WP1220, WP1219,
      WP16,
      WP60, WP61, WP63, WP65, WP67,
      WP68, WP69, WP70, WP71, WP74
    )
  )
  # Get rid of SPSS attributes
  gwp <- gwp %>%
    zap_labels() %>%
    zap_label() %>%
    zap_widths() %>%
    zap_formats()
  # Save into a good format
  write_rds(gwp, gwp_path)
} else {
  gwp <- read_rds(gwp_path)
}

# Rename and recode variables
gwp <- gwp %>%
  clean_names() %>%
  transmute(
    country = countrynew,
    year = year_calendar,
    id = wpid,
    sex = factor(wp1219, levels = c(2, 1), labels = c("Female", "Male")),
    age = wp1220,
    # Don't know, refused, and missing values
    Life_satisfaction = if_else(between(wp16, 0, 10), wp16, NaN),
    across(wp60:wp74, ~ if_else(between(., 1, 2), ., NaN)),
    # Also reverse the weird 1: yes 2: no coding here
    across(wp60:wp74, ~ 3 - .)
  )

# Scale scores note rescaling to percentages
gwp <- gwp %>%
  mutate(
    Life_satisfaction = Life_satisfaction * 10,
    Negative_experiences =
      (rowMeans(select(., wp68:wp74), na.rm = TRUE) - 1) * 100,
    Positive_experiences =
      (rowMeans(select(., wp60:wp67), na.rm = TRUE) - 1) * 100
  ) %>%
  select(-c(wp60:wp74))

# Categorize ages
gwp <- filter(gwp, age <= 89)
gwp <- gwp %>%
  mutate(
    age = cut(
      age,
      breaks = seq(15, 90, by = 5),
      include.lowest = TRUE,
      right = FALSE,
      labels = paste(seq(15, 85, by = 5), "to", seq(19, 89, by = 5))
    ) %>% as.character()
  )

# There is one 13 year old which resulted in a NA and we drop it
gwp <- drop_na(gwp, age)

# Some sex values are missing, drop those
gwp <- drop_na(gwp, sex)

# Pivot outcomes to long format
gwp <- gwp %>%
  pivot_longer(
    c(
      Life_satisfaction,
      Negative_experiences,
      Positive_experiences
    ),
    names_to = "outcome", values_to = "val"
  )

# Aggregate outcomes
gwp <- gwp %>% 
  summarise(
    n = n(),
    sd = sd(val, na.rm = TRUE),
    val = mean(val, na.rm = TRUE),
    .by = c(country, year, sex, age, outcome)
  )

# Assign N of 1 cells a standard deviation of zero
gwp <- gwp %>% 
  mutate(sd = if_else(is.na(sd), 0, sd))

# We can see that this would result in North Cyprus being lumped with Cyprus so we need to replace the name before harmonising

# Harmonise old names and replace only if harmonised name found
gwp <- gwp %>%
  mutate(
    country_harmonised = countryname(
      country,
      destination = "cldr.short.en"
    )
  ) %>%
  # This picks the harmonised name if exists, otherwise original name
  # Prevent north cyprus from becoming cyprus
  mutate(
    country_harmonised = ifelse(
      country == "Northern Cyprus",
      "Northern Cyprus",
      country_harmonised
    )
  ) %>%
  mutate(country = coalesce(country_harmonised, country)) %>%
  select(-country_harmonised)

gwp_mock_path <- "data-raw/gwp-MOCK.rds"
if (dir.exists("data-raw/Gallup") & !file.exists(gwp_mock_path)) {
  library(synthpop)
  gwp_syn <- gwp %>% 
    drop_na()
  gwp_syn <- syn(
    gwp_syn, 
    method = c("", "", "", "", "", "", "normrank", "normrank"),
    maxfaclevels = 168
  )
  gwp_syn <- tibble(gwp_syn$syn)
  saveRDS(gwp_syn, gwp_mock_path)
}

# Load if actual data doesn't exist
if (!dir.exists("data-raw/Gallup")) {
  cat("Using synthetic GWP data.")
  gwp <- read_rds(gwp_mock_path)
} else {
  cat("Using actual GWP data.")
}

# Load data files and merge to one table
gbd <- list.files(
  "data-raw/GBD/",
  pattern = ".zip", full.names = TRUE, recursive = TRUE
) %>%
  read_csv()

# First step is to clean names and remove unnecessary variables
gbd <- clean_names(gbd)
# distinct(gbd, measure, metric)
gbd <- select(gbd, -measure, -metric)

# Rates to percentages
gbd <- gbd %>%
  mutate(across(c(val, upper, lower), ~ .x / 1000))

# Clean cause names
# distinct(gbd, cause)
gbd <- gbd %>%
  mutate(
    cause = case_when(
      cause == "Anxiety disorders" ~ "Anxiety",
      cause == "Depressive disorders" ~ "Depression",
      cause == "Self-harm" ~ "Selfharm"
    )
  )

# Harmonise variable names with other datasets
gbd <- gbd %>%
  rename(country = location, outcome = cause)

# The outcomes values are model predictions and come with lower and upper CI limits (2.5 and 97.5 %iles of their posterior distributions). We convert those to normal approximate standard errors.
gbd <- gbd %>%
  mutate(sd = (upper - lower) / (1.96 * 2)) %>%
  select(-c(upper, lower))

# Harmonise old names and replace only if harmonised name found
gbd <- gbd %>%
  mutate(
    country_harmonised = countryname(
      country,
      destination = "cldr.short.en"
    )
  ) %>%
  # This picks the harmonised name if exists, otherwise original name
  mutate(country = coalesce(country_harmonised, country)) %>%
  select(-country_harmonised)

# Download data sets from the ITU website if not yet downloaded
if (!file.exists("data-raw/ITU/PercentIndividualsUsingInternet.xlsx")) {
  dir.create("data-raw/ITU")
  download.file(
    "https://www.itu.int/en/ITU-D/Statistics/Documents/statistics/2022/July/PercentIndividualsUsingInternet.xlsx",
    "data-raw/ITU/PercentIndividualsUsingInternet.xlsx"
  )
  download.file(
    "https://www.itu.int/en/ITU-D/Statistics/Documents/statistics/2022/July/MobileBroadbandSubscriptions_2007-2021.xlsx",
    "data-raw/ITU/MobileBroadbandSubscriptions_2007-2020.xlsx"
  )
}

# Read and reshape mobile broadband data
itu_mobile <- read_xlsx(
  "data-raw/ITU/MobileBroadbandSubscriptions_2007-2020.xlsx",
  sheet = 2,
  col_types = "text"
) %>%
  pivot_longer(-c(Indicator, Economy)) %>%
  separate(name, into = c("year", "variable")) %>%
  pivot_wider(names_from = variable) %>%
  mutate(mobile = as.numeric(value)) %>%
  select(country = Economy, year, mobile)

# Percent using internet
itu_percent <- read_xlsx(
  "data-raw/ITU/PercentIndividualsUsingInternet.xlsx",
  col_types = "text"
) %>%
  pivot_longer(-c(Indicator, Economy)) %>%
  separate(name, into = c("year", "variable")) %>%
  pivot_wider(names_from = variable) %>%
  mutate(internet = as.numeric(value)) %>%
  select(country = Economy, year, internet)

# Merge to one table in wide format (oldest on left to include all years)
itu <- reduce(list(itu_percent, itu_mobile), left_join)

# years to numbers
itu <- itu %>%
  mutate(year = as.numeric(year))

# Drop tables
rm(itu_mobile, itu_percent)

# Harmonise old names and replace only if harmonised name found
itu <- itu %>%
  mutate(
    country_harmonised = countryname(
      country,
      destination = "cldr.short.en"
    )
  ) %>%
  # This picks the harmonised name if exists, otherwise original name
  mutate(country = coalesce(country_harmonised, country)) %>%
  select(-country_harmonised)

# Drop countries that don't exist in outcomes
itu <- itu %>%
  filter(
    country %in% unique(gbd$country) | country %in% unique(gwp$country)
  )

# Drop countries that don't have any actual internet data
itu_zeros <- itu %>%
  group_by(country) %>%
  summarise(s = sum(internet, na.rm = TRUE)) %>%
  filter(s == 0)

itu <- itu %>%
  filter(!(country %in% itu_zeros$country))
rm(itu_zeros)

# Reshape to easily impute all variables
itu_long <- itu %>%
  pivot_longer(c(internet, mobile))

# Impute by country and variable
itu_long <- itu_long %>%
  group_by(country, name) %>%
  # Find first and last year with this variable in each country
  # so as to prevent extrapolation
  mutate(
    min_year = year[min(which(!is.na(value)))],
    max_year = year[max(which(!is.na(value)))]
  ) %>%
  # Interpolate values linearly
  # use possibly() to avoid errors for countries with no data
  mutate(
    across(
      value,
      .fns = list(
        i = possibly(
          function(x) na_interpolation(x, maxgap = Inf),
          otherwise = NaN
        )
      )
    )
  ) %>%
  # Take out projected (non-intermediate) imputed values
  mutate(
    value_i = if_else(
      between(year, unique(min_year), unique(max_year)),
      value_i,
      NaN
    )
  ) %>%
  ungroup() %>%
  select(-min_year, -max_year)

# Summary
itu_sum <- itu_long %>%
  filter(!(name=="mobile" & year < 2007)) %>% 
  group_by(Metric = name) %>%
  summarise(
    total = n(), 
    total_not_missing = sum(!is.na(value)),
    missing = sum(is.na(value)),
    imputed = sum(is.na(value) & !is.na(value_i))
  )

# We then replace the original values with the imputed ones
itu <- itu_long %>%
  # Pick non-missing value, from original and imputed
  mutate(value = coalesce(value, value_i)) %>%
  select(-value_i) %>%
  pivot_wider()
rm(itu_long)

regions <- bind_rows(
  distinct(itu, country),
  distinct(gbd, country),
  distinct(gwp, country)
) %>% 
  distinct() %>% 
  mutate(
    region = countrycode(
      country, 
      origin = "country.name", 
      destination = "continent"
    )
  )

# Manually fix two countries
regions <- regions %>% 
  mutate(
    region = if_else(country == "Kosovo", "Europe", region),
    region = if_else(country == "Nagorno Karabakh", "Asia", region)
  )

# Add regions to data tables
itu <- left_join(itu, regions)
gbd <- left_join(gbd, regions)
gwp <- left_join(gwp, regions)

# Stack outcome data
dat <- bind_rows(gwp, gbd)

# Merge predictors to outcomes
dat <- left_join(dat, itu)

# Center year
dat <- dat %>%
  mutate(year = (year - 2010))

# Ensure that categorical variables are factors
dat <- dat %>%
  mutate(across(c(sex, age), factor))

# Arrange on outcome
oo <- c(
  "Life_satisfaction", "Negative_experiences", "Positive_experiences",
  "Anxiety", "Depression", "Selfharm"
)
dat <- dat %>%
  mutate(outcome = factor(outcome, levels = oo)) %>% 
  arrange(outcome, country, region, year)

# Save data for models
saveRDS(dat, "data/data-all.rds")
```

In 2005, an estimated 17% of the global population used the internet, but by 2020 this number was already 59% [@ituITUICTStatistics2021]. Accompanying the rapid spread of the internet, worries have proliferated that its broad adoption—and technologies enabled by it such as online games, smartphones, and social media—is actively harming its users, particularly adolescents [@carrShallowsHowInternet2010; @turkleAloneTogetherWhy2011]. As a response, users and a growing number of national governments have acted in limiting access to online technologies [@buckleyChinaTightensLimits2021; @departmentfordigitalculturemedia&sportDraftOnlineSafety2021; @kattulaGamePoliciesContexts2021].

However, evidence for widespread harms of online technologies is limited. Initial reports of internet-facilitated harms have been challenged by later work informed by many methodologies including longitudinal [@jensenYoungAdolescentsDigital2019], specification curve analyses [@orbenAssociationAdolescentWellbeing2019], meta-analyses [@appelAreSocialMedia2020], and systematic reviews [@bestOnlineCommunicationSocial2014; @dickson2018screen; @odgersAnnualResearchReview2020; @ophirNewMediaScreenTime2020]. Reviews indicate early research has been hampered by inaccurate measurements of engagement with internet and related technologies [@davidsonFuzzyConstructsTechnology2022; @parrySystematicReviewMetaanalysis2021; @scharkowAccuracySelfReportedInternet2016], biased convenience samples drawn predominantly from countries in the Global North [ghaiLackSampleDiversity2023; @ghaiSocialMediaAdolescent2022], studying a limited range of well-being outcomes [@orbenSocialMediaEnduring2019], and reliance on self-reported evaluations in place of clinical estimates of important mental health outcomes [@campbellInternetUseSocially2006]. A comprehensive test of the overall association between internet adoption and well-being and mental health, broadly defined, has therefore not been conducted. As a consequence, much of the evidence purporting to show that internet technology adoption is associated with broad changes in well-being and mental health, or even causes negative outcomes, remains equivocal [@dickson2018screen; @hawkesCMOReportUnable2019].

@@

```{r gwp-data-summaries}
gwp_sum <- gwp %>% 
  group_by(outcome) %>% 
  summarise(n = sum(n), cells = n()) %>% 
  mutate(across(n:cells, ~number(., big.mark = ",")))
```

@@

We identified the International Telecommunication Union's (ITU) database of information and communications technology (ICT) as the most comprehensive country-level source of time-series data on internet adoption [@ituITUICTStatistics2021]. The ITU has collated, from 222 countries' statistical and telecommunications agencies, the yearly percentages of population using the internet from 2000 to `r max(itu$year)`, and yearly per capita broadband subscriptions from 2007 to `r max(itu$year)`. Our analyses used `r number(itu_sum[1, 3, drop = TRUE], big.mark = ",")` yearly internet percentages and `r number(itu_sum[2, 3, drop = TRUE], big.mark = ",")` broadband subscription rates; we imputed `r number(itu_sum[1, 5, drop = TRUE], big.mark = ",")` (`r number(itu_sum[2, 5, drop = TRUE], big.mark = ",")` mobile) intermediate missing values to this dataset using linear interpolation without extrapolation.

@@

We examined subjective well-being indicators from the Gallup World Poll (GWP), a nationally representative annual survey of 1,000 civilian, non-institutionalised individuals aged 15 years or older from `r length(unique(gwp$country))` countries from 2005 to 2021. The surveys are conducted face-to-face or via telephone, in the respondents' native language and by local interviewers. For details on the GWP sampling and survey methodology, see [@gallupGallupWorldPoll2020; @gallupHowDoesGallup2014].

@@

We analysed the data with meta-analytic bayesian hierarchical regression models [@burknerBrmsPackageBayesian2017; @gelmanDataAnalysisUsing2007; @standevelopmentteamStanModelingLanguage2021]. We first asked how the well-being outcomes had changed over time. Bayesian methods are especially suitable to our reserch questions, because they involve many potential contrasts between e.g. countries and ages. Using bayesian hierarchical models allowed us to accurately quantify uncertainty at different levels of analyses while avoiding problems with multiple comparisons (see below). To answer, we modelled each outcome (e.g. life satisfaction) on population level intercepts (we centred time on 2010), contrasts of time (in continuous years), sex (Female: 1, Male: -1), and their interaction. We allowed all coefficients to vary across countries, age groups, and the age by country interactions, and treated all outcomes as normally distributed.

@@

We specified the model as

\begin{align*} 
y_i &\sim \text{Normal}(\mu_i, \sqrt{\sigma^{2} + \text{sd}^2_i}), \\
\mu_i &= \alpha_{0} + \beta_{0\text{country}[i]} + \gamma_{0\text{age}[i]} + \delta_{0\text{age:country}[i]} + \\
&\quad\ (\alpha_{1} + \beta_{1\text{country}[i]} + \gamma_{1\text{age}[i]} + \delta_{1\text{age:country}[i]})\text{Time}_i + \\
&\quad\ (\alpha_{2} + \beta_{2\text{country}[i]} + \gamma_{2\text{age}[i]} + \delta_{2\text{age:country}[i]})\text{Sex}_i + \\
&\quad\ (\alpha_{3} + \beta_{3\text{country}[i]} + \gamma_{3\text{age}[i]} + \delta_{3\text{age:country}[i]})\text{Sex}_i \times \text{Time}_i + \\
&\quad\ (\alpha_{4} + \beta_{4\text{country}[i]} + \gamma_{4\text{age}[i]} + \delta_{4\text{age:country}[i]})\text{X}^{\text{CW}}_i + \\
&\quad\ (\alpha_{5} + \beta_{5\text{country}[i]} + \gamma_{5\text{age}[i]} + \delta_{5\text{age:country}[i]})\text{Sex}_i \times \text{X}^{\text{CW}}_i + \\
&\quad\ \alpha_6 \text{X}^{\text{CB}}_i,
\\
\pmb{\beta} &\sim \text{MVN}(\pmb{0}, \Sigma^\text{country}), \\
\pmb{\gamma} &\sim \text{MVN}(\pmb{0}, \Sigma^\text{age}), \\
\pmb{\delta} &\sim \text{MVN}(\pmb{0}, \Sigma^\text{age:country}),
\end{align*}

where $i$ indexes rows in the data and $\text{sd}_i$ are the known standard deviations. $\text{X}^{\text{CW}}_i$ and $\text{X}^{\text{CB}}_i$ are the within- and between-country centred internet technology adoption predictor (per capita internet users or mobile broadband subscriptions, in separate models), respectively. In addition to the standard deviations, we weighted each observation by the number of respondents in that cell, to adjust for varying sample sizes across cells. We used default noninformative priors on all parameters.

@@

We conducted all data analyses with R [@rcoreteamLanguageEnvironmentStatistical2021] and estimated the models using Stan's Hamiltonian Monte Carlo sampling via the brms R package [@burknerBrmsPackageBayesian2017; @standevelopmentteamStanModelingLanguage2021]. We ran 4 HMC chains with 4,000 iterations and first 2,000 as warmup for 8,000 total iterations. We ensured that all population-level parameters' had converged using numeric diagnostics (all Rhats were within 0.01 of 1) and visual inspection of parameters' scatterplot matrices.

@@

```{r data-describe}
gwp %>% 
  group_by(outcome) %>% 
  summarise(n=sum(n)) %>% 
  summarise(N=comma(median(n))) %>% 
  pull(N)

bind_rows(
  "GWP" = gwp,
  "GBD" = gbd,
  .id = "Dataset"
) %>% 
  mutate(Dataset = fct_inorder(Dataset)) %>% 
  count(Dataset, outcome) %>% 
  bind_rows(
    itu %>% 
      select(internet, mobile) %>% 
      pivot_longer(everything(), names_to = "outcome") %>% 
      drop_na(value) %>% 
      count(outcome) %>% 
      mutate(Dataset = "ITU")
  )

# Individuals using the internet per 100 inhabitants | World
if (!file.exists(
  "data-raw/ITU/ITU_regional_global_Key_ICT_indicator_aggregates_rev1_Jan_2022.xlsx")) {
  download.file("https://www.itu.int/en/ITU-D/Statistics/Documents/facts/ITU_regional_global_Key_ICT_indicator_aggregates_rev1_Jan_2022.xlsx", "data-raw/ITU/ITU_regional_global_Key_ICT_indicator_aggregates_rev1_Jan_2022.xlsx")
}
read_excel("data-raw/ITU/ITU_regional_global_Key_ICT_indicator_aggregates_rev1_Jan_2022.xlsx", range = "T85:AI86")
```

```{r models-load}
# Create a table of outcomes in appropriate order, so can map functions over this
oo <- c(
  "Life_satisfaction", "Negative_experiences", "Positive_experiences",
  "Anxiety", "Depression", "Selfharm"
)

>>>>

---
title: Global well-being and mental health in the internet age
shorttitle: Global well-being
leftheader: Global well-being
author: 
  - name: Matti Vuorre
    affiliation: "1,2"
    address: Tilburg University
    email: m.j.vuorre@tilburguniversity.edu
  - name: Andrew K. Przybylski
    affiliation: "2"
    corresponding: yes
    address: Oxford Internet Institute
    email: andy.przybylski@oii.ox.ac.uk
affiliation:
  - id: 1
    institution: Tilburg School of Social and Behavioral Sciences, Tilburg University
  - id: 2
    institution: Oxford Internet Institute, University of Oxford
authornote: |
  \noindent \textbf{This pre-print is not yet peer-reviewed.}
abstract: | 
  In the last two decades the widespread adoption of internet technologies has inspired concern that they have negatively impacted mental health and psychological well-being. However, research on the topic is contested and hampered by methodological shortcomings leaving the broader consequences of internet adoption unknown. We show that the past two decades have seen only small and inconsistent changes in global well-being and mental health that are not suggestive of the idea that the adoption of internet and mobile broadband is consistently linked to negative psychological outcomes. Further investigation of this topic requires transparent study of online behaviours where they occur, namely on online platforms. We call for increased collaborative efforts between independent scientists and the internet technology sector.

keywords: well-being, mental health, internet technology, technology effects
wordcount: '`r scales::number(wordcountaddin:::word_count("ms.Rmd"), big.mark = ",")`'
bibliography: "references.bib"
floatsintext: yes
linenumbers: no
draft: no
mask: no
figurelist: no
tablelist: no
footnotelist: no
csl: "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
documentclass: apa7
classoption: jou
output: 
  papaja::apa6_pdf:
    number_sections: false
    keep_tex: true
  papaja::apa6_docx: 
    number_sections: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
# Packages
library(papaja)
library(scales)
library(tidybayes)
library(ggdist)
library(lubridate)
library(brms)
library(patchwork)
library(knitr)
library(posterior)
library(bayestestR)
library(tidyverse)

opts_chunk$set(
  eval = TRUE,
  cache = TRUE,
  include = FALSE,
  message = FALSE,
  warning = FALSE,
  error = TRUE,
  dpi = 300,
  fig.align = "center"
)

# Plotting options
theme_set(
  theme_linedraw(
    base_size = if_else(knitr::is_latex_output(), 8, 7)
  ) +
    theme(
      panel.grid = element_blank(),
      strip.text = element_text(margin = margin(4, 4, 4, 4, "pt")),
      axis.text = element_text(size = rel(0.7)),
      axis.ticks = element_line(size = rel(0.2))
    )
)
dir.create("cache", FALSE)
```

```{r data}
dat <- read_rds("data/data-all.rds")
gwp <- read_rds("data/gwp.rds")
itu <- read_rds("data/itu.rds")
gbd <- read_rds("data/gbd.rds")
z <- read_rds("data/scaling-factors.rds")
coefficients <- read_rds("cache/coefficients.rds")
epred <- read_rds("cache/epred.rds")
```

```{r variables}
# For ordering outcome factor
oo <- c(
  "Life_satisfaction", 
  "Negative_experiences", 
  "Positive_experiences",
  "Anxiety", "Depression", "Selfharm"
)
```

@@

In 2005, an estimated 17% of the global population used the internet, but by 2020 this number was already 59% [@ituITUICTStatistics2021]. Accompanying the rapid spread of the internet, worries have proliferated that its broad adoption—and technologies enabled by it such as online games, smartphones, and social media—is actively harming its users, particularly adolescents [@carrShallowsHowInternet2010; @turkleAloneTogetherWhy2011]. As a response, users and a growing number of national governments have acted in limiting access to online technologies [@buckleyChinaTightensLimits2021; @departmentfordigitalculturemediasportDraftOnlineSafety2021; @kattulaGamePoliciesContexts2021].

However, evidence for widespread harms of online technologies is limited. Initial reports of internet-facilitated harms have been challenged by later work informed by many methodologies including longitudinal [@jensenYoungAdolescentsDigital2019], specification curve analyses [@orbenAssociationAdolescentWellbeing2019], meta-analyses [@appelAreSocialMedia2020], and systematic reviews [@bestOnlineCommunicationSocial2014; @dickson2018screen; @odgersAnnualResearchReview2020; @ophirNewMediaScreenTime2020]. Reviews indicate early research has been hampered by inaccurate measurements of engagement with internet and related technologies [@davidsonFuzzyConstructsTechnology2022; @parrySystematicReviewMetaanalysis2021; @scharkowAccuracySelfReportedInternet2016], biased convenience samples drawn predominantly from countries in the Global North [@ghaiLackSampleDiversity2023; @ghaiSocialMediaAdolescent2022], studying a limited range of well-being outcomes [@orbenSocialMediaEnduring2019], and reliance on self-reported evaluations in place of clinical estimates of important mental health outcomes [@campbellInternetUseSocially2006]. A comprehensive test of the overall association between internet adoption and well-being and mental health, broadly defined, has therefore not been conducted. As a consequence, much of the evidence purporting to show that internet technology adoption is associated with broad changes in well-being and mental health, or even causes negative outcomes, remains equivocal [@dickson2018screen; @hawkesCMOReportUnable2019].

@@

```{r data-summary-wb}
gwp_sum <- gwp %>% 
  group_by(outcome) %>% 
  summarise(n = sum(n), cells = n()) %>% 
  mutate(across(n:cells, ~number(., big.mark = ",")))
gwp_sum
```

@@

```{r data-summary-itu}
itu %>% 
  summarise(
    i = sum(!is.na(internet)),
    m = sum(!is.na(mobile))
  )
```

We identified the International Telecommunication Union's (ITU) database of information and communications technology (ICT) as the most comprehensive country-level source of time-series data on internet adoption [@ituITUICTStatistics2021]. The ITU has collated, from 222 countries' statistical and telecommunications agencies, the yearly percentages of population using the internet from 2000 to 2021, and yearly per capita mobile broadband subscriptions from 2007 to 2021. Our analyses used 4,200 yearly internet percentages and 2,466 broadband subscription rates.

@@

We examined subjective well-being indicators from the Gallup World Poll (GWP), a nationally representative annual survey of 1,000 civilian, non-institutionalised individuals aged 15 years or older from `r length(unique(gwp$country))` countries from 2005 to 2022. The surveys are conducted face-to-face or via telephone, in the respondents' native language and by local interviewers. For details on the GWP sampling and survey methodology, see [@gallupGallupWorldPoll2020; @gallupHowDoesGallup2014].

@@

We analysed the data with meta-analytic bayesian hierarchical regression models [@burknerBrmsPackageBayesian2017; @gelmanDataAnalysisUsing2007; @standevelopmentteamStanModelingLanguage2022]. Bayesian methods are especially suitable to our reserch questions, because they involve many potential contrasts between e.g. countries and ages. Using bayesian hierarchical models allowed us to accurately quantify uncertainty at different levels of analyses while avoiding problems with multiple comparisons (see below).

We first asked how the well-being outcomes had changed over time. To answer, we modelled each outcome (e.g. life satisfaction) on population level intercepts (we centred time on 2010), contrasts of time (in continuous years), sex (Male: -1, Female: 1), and their interaction. We allowed all coefficients to vary across countries, age groups, and the age by country interactions, and treated all outcomes as normally distributed.

@@

More formally, we specified the model as

\begin{align*} 
y_i &\sim \text{Normal}(\mu_i, \sqrt{\sigma^{2} + \text{se}^2_i}), \\
\mu_i &= \alpha_{0} + \beta_{0\text{country}[i]} + \gamma_{0\text{age}[i]} + \delta_{0\text{age:country}[i]} + \\
&\quad\ (\alpha_{1} + \beta_{1\text{country}[i]} + \gamma_{1\text{age}[i]} + \delta_{1\text{age:country}[i]})\text{Time}_i + \\
&\quad\ (\alpha_{2} + \beta_{2\text{country}[i]} + \gamma_{2\text{age}[i]} + \delta_{2\text{age:country}[i]})\text{Sex}_i + \\
&\quad\ (\alpha_{3} + \beta_{3\text{country}[i]} + \gamma_{3\text{age}[i]} + \delta_{3\text{age:country}[i]})\text{Sex}_i \times \text{Time}_i + \\
&\quad\ (\alpha_{4} + \beta_{4\text{country}[i]} + \gamma_{4\text{age}[i]} + \delta_{4\text{age:country}[i]})\text{X}^{\text{CW}}_i + \\
&\quad\ (\alpha_{5} + \beta_{5\text{country}[i]} + \gamma_{5\text{age}[i]} + \delta_{5\text{age:country}[i]})\text{Sex}_i \times \text{X}^{\text{CW}}_i + \\
&\quad\ \alpha_6 \text{X}^{\text{CB}}_i,
\\
\pmb{\beta} &\sim \text{MVN}(\pmb{0}, \Sigma^\text{country}), \\
\pmb{\gamma} &\sim \text{MVN}(\pmb{0}, \Sigma^\text{age}), \\
\pmb{\delta} &\sim \text{MVN}(\pmb{0}, \Sigma^\text{age:country}),
\end{align*}

where $i$ indexes rows in the data and $\text{se}_i$ are the known standard errors. $\text{X}^{\text{CW}}_i$ and $\text{X}^{\text{CB}}_i$ are the within- and between-country centred internet technology adoption predictor (per capita internet users or mobile broadband subscriptions, in separate models), respectively. We used default noninformative priors [@burknerBrmsPackageBayesian2017] on all parameters except the following population-level coefficients, that we modelled as

\begin{align*}
\alpha_{1,3-6} &\sim \text{Normal}(0, 1), \\
\alpha_{2} &\sim \text{Normal}(0, 3).
\end{align*}

We used these priors to help model convergence, and on the assumption that differences in well-being between males and females are likely greater than differences associated with years, one-percent increases in technology adoption predictors, or the interaction terms.

@@

We conducted all data analyses with R [@rcoreteamLanguageEnvironmentStatistical2022] and estimated the models using Stan's Hamiltonian Monte Carlo sampling via the brms R package [@burknerBrmsPackageBayesian2017; @standevelopmentteamStanModelingLanguage2022].

Our main descriptive focus was on raw regression coefficients, which indicate contrasts in the outcomes on the percentage scale [@cohenProblemUnitsCircumstance1999] as a function of one unit (year / percent) increase in the predictor. We report these parameters' 95% Credibility Intervals (CI) and posterior probabilities of direction ($p^d$). The latter indicate the degree of certainty in the observed direction of a coefficient, and are sometimes numerically similar to one minus a one-sided *p*-value.

In addition, on a reviewer's request we report standardized coefficients to describe the magnitudes in an alternative context. These standardized coefficients indicate contrasts in the z-scored outcomes (life satisfaction SD: `r number(z$val_sd[1], .1)`; negative experiences: `r number(z$val_sd[4], .1)`; positive experiences: `r number(z$val_sd[7], .1)`) as a function of the average year-on-year changes in internet (`r number(z$xdev[2], .1)`%) or mobile broadband (`r number(z$xdev[3], .1)`%) adoption. To determine the practical significance of the standardized associations, we report posterior probabilities that their magnitudes exceed the region of practical equivalence to zero ($p^{\text{ROPE}}$) [@kruschkeRejectingAcceptingParameter2018] as defined by [-0.1, 0.1] [@fergusonProvidingLowerboundEstimate2021].

@@

Figure \@ref(fig:fig-means-wb)A describes the levels of internet and mobile broadband adoption, and the three well-being variables over time averaged over sex and age. Figure \@ref(fig:fig-means-wb)A shows the universal global penetration of internet technologies in the past two decades, and suggests that in contrast, changes in well-being are likely to be small.

```{r tbl-avg-wb}
#| include: true

tbl_average <- coefficients %>% 
  filter(
    level == "average",
    sex == "average"
  ) %>% 
  select(
    Predictor = predictor, 
    Outcome = outcome, 
    `Raw estimate` = Raw, 
    `Standardized estimate` = Scaled
  )

tbl_average %>% 
  filter(Outcome %in% oo[1:3]) %>% 
  mutate(
    Predictor = if_else(Outcome %in% oo[c(1, 4)], Predictor, ''),
    Outcome = str_replace(Outcome, "_", " ")
  ) %>% 
  # Justify considering negative estimates
  mutate(
    across(
      3:4, 
      ~if_else(
        str_starts(., "-"), 
        ., 
        paste0("\\phantom{-}", .)
      )
    )
  ) %>% 
  apa_table(
    caption = "Well-being: Population-level estimates",
    span_text_columns = TRUE, 
    font_size = "small",
    escape = FALSE,
    note = "Raw estimates are contrasts in outcome percentages associated with a one year or percentage increase in the predictor. Numbers indicate posterior means, [95\\%CIs], and ($p^d$; posterior probabilities of direction). Standardized estimates refer to contrasts in the z-scored outcome as function of the average year-on-year change in the predictor. The latter percentages indicate posterior probabilities inside a region of practical equivalence to zero ([-0.1, 0.1]; $p^{\\text{ROPE}}$)."
  )
```

### Trends
